Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-23 appendix.tex
fig updates, notation updates except for sec 3

\section{Full governing equation for vertically isothermal disks in
  the vertically global shearing box}\label{adia_improve}
Here we explain the derivation of Eq. \ref{nearly_iso_explicit} used in our 
analytical discussion in \S\ref{analytical}. We shall need the
vertical derivative of the epicyclic frequency $\kappa$. In the global
disk with $\Gamma=1$ we have 
\begin{align}\label{dkappa2}
  \frac{\p\kappa^2}{\p z} = 4 \frac{\p\Omega^2}{\p z} + r\frac{\p}{\p
    r}\frac{\p\Omega^2}{\p z} = -
  \frac{\p\ln\rho}{\p z}\frac{qc_s^2}{r^2} \left(
    \frac{3z^2}{r^2+z^2}-1\right) \equiv - \frac{\p\ln\rho}{\p z}\frac{qc_s^2}{r^2}F(r,z).
\end{align}
% For the local problem we may then write
% \begin{align}
%   \frac{dD}{dz}  = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2}F(z),
% \end{align}
The function $F$ increases monotonically from $F=-1$ at $z=0$ to $F\to2$
as $|z|\to\infty $, so $|F|=O(1)$.

@@

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
A limitation of the VGSB is that it cannot be used for 
adiabatic flow, even if the above conditions are satisfied. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ in 
Eq. \ref{vertiso_gov_nondim} % (which itself is obtained by setting
% $\hat{g}_c=0$ in Eq. \ref{lin_mass}---\ref{lin_energy})
to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii h q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\upsilon}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii h  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\upsilon}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii h
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii h q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\upsilon}^2$ is real, and  $\hat{\upsilon}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\upsilon}^2$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability for adiabatic
perturbations in a thin, vertically isothermal disk would require 
$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the global radial
disk structure responsible for vertical shear
($\p_z\Omega$), but ignores it elsewhere in the linear problem. 
Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI because it occurs
for $\beta\ll1$.   %and small scale

@@

We seek power series solutions to Eq. \ref{iso_ode3}, 
\begin{align}
  W(\zhat) = \sum_{l=0}^\infty a_l\zhat^l. 
\end{align}
Then the coefficients $a_l$ are given by the recurrence relation
\begin{align}
  (l+2)(l+1)a_{l+2} +
  \left[\hat{\upsilon}^2\left(1+\frac{\khat^2}{1-\hat{\upsilon}^2}\right)
    - l\left(1+\frac{\ii h q
        \hat{k}}{1-\hat{\upsilon}^2}\right)\right] a_l = 0. 
\end{align}
We demand the series to terminate  at $l=L$, i.e. a polynomial of
order $L$, so that the vertical kinetic energy remain bounded as 
$|\zhat|\to\infty$.  Then the eigenfrequency $\hat{\upsilon}$ is given via 
\begin{align}
\hat{\upsilon}^4 - \left(L+1+\khat^2\right)\hat{\upsilon}^2 + L\left(1 +
  \ii h q \khat\right) = 0. \label{simple_growth}
\end{align}

>>>>

\section{Full governing equation for vertically isothermal disks in
  the vertically global shearing box}\label{adia_improve}
Here we explain the derivation of Eq. \ref{nearly_iso_explicit} used in our 
analytical discussion in \S\ref{analytical}. We shall need the
vertical derivative of the epicyclic frequency $\kappa$. In the
vertically isothermal global
disk with $\Gamma=1$ we have 
\begin{align}\label{dkappa2}
  \frac{\p\kappa^2}{\p z} = 4 \frac{\p\Omega^2}{\p z} + r\frac{\p}{\p
    r}\frac{\p\Omega^2}{\p z} = -
  \frac{\p\ln\rho}{\p z}\frac{qc_s^2}{r^2} \left(
    \frac{3z^2}{r^2+z^2}-1\right) \equiv - \frac{\p\ln\rho}{\p z}\frac{qc_s^2}{r^2}F(r,z).
\end{align}
% For the local problem we may then write
% \begin{align}
%   \frac{dD}{dz}  = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2}F(z),
% \end{align}
The function $F$ increases monotonically from $F=-1$ at $z=0$ to $F\to2$
as $|z|\to\infty $, so $|F|=O(1)$.

@@

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
A limitation of the VGSB is that it cannot be used for 
adiabatic flow, even if the above conditions are satisfied. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ and hence $\chi = 1/\gamma$ in 
Eq. \ref{vertiso_gov_nondim} % (which itself is obtained by setting
% $\hat{g}_c=0$ in Eq. \ref{lin_mass}---\ref{lin_energy})
to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii h q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\upsilon}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii h  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\upsilon}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii h
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii h q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\upsilon}^2$ is real, and  $\hat{\upsilon}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\upsilon}^2$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability for adiabatic
perturbations in a thin, vertically isothermal disk would require 
$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the global radial
disk structure responsible for vertical shear
($\p_z\Omega$), but ignores it elsewhere in the linear problem. 
Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI, which occurs 
for $\beta\ll1$.   %and small scale

@@

We seek power series solutions to Eq. \ref{iso_ode3}, 
\begin{align}
  W(\zhat) = \sum_{l=0}^\infty a_l\zhat^l. 
\end{align}
Then the coefficients $a_l$ are given by the recurrence relation
\begin{align}
  (l+2)(l+1)a_{l+2} +
  \left[\hat{\upsilon}^2\left(1+\frac{\khat^2}{1-\hat{\upsilon}^2}\right)
    - l\left(1+\frac{\ii h q
        \hat{k}}{1-\hat{\upsilon}^2}\right)\right] a_l = 0. 
\end{align}
We demand the series to terminate  at $l=L$, i.e. a polynomial of
order $L$, so that the vertical kinetic energy density remain bounded as 
$|\zhat|\to\infty$.  Then the eigenfrequency $\hat{\upsilon}$ is given via 
\begin{align}
\hat{\upsilon}^4 - \left(L+1+\khat^2\right)\hat{\upsilon}^2 + L\left(1 +
  \ii h q \khat\right) = 0. \label{simple_growth}
\end{align}

### 2015-04-23 linear_iso.tex
fig updates, notation updates except for sec 3

In preparation for a numerical study of the above equations, we will 
first present a fully analytic discussion for vertically isothermal
disks, for which $\Gamma=1$ and $c_s$ is vertically constant. To this end, 
we introduce the following simplifications:
%In Appendix \ref{adia_improve} we derive  a with the following simplifications:

@@

\item We consider eigenvalues such that
  $|\upsilon^2|\ll \kappa^2,\,\Omega^2$. We call this the low frequency 
  approximation

\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$ for vertically isothermal disks. This is already captured
  explicitly in the linearized equations through the term
  $d\Omega/dz$. We thus ignore the vertical dependence of $\Omega$ and
  $\kappa$ elsewhere, and replace them with
  $\Omega_K$.  We call this the Keplerian approximation.

@@

 In Appendix \ref{global_corr}  we discuss limitations of using
  the VGSB, where we show that this approximation is valid provided we 
  consider consider fast thermal relaxation times
  ($\beta\lesssim 1$) and perturbations with small radial lengthscales
  ($\khat\gtrsim 1$), as appropriate for the VSI.

>>>>

In preparation for a numerical study of the above equations, we will 
first give a fully analytic discussion for vertically isothermal
disks, for which $\Gamma=1$ and $c_s$ is vertically constant. To this end, 
we introduce the following simplifications:
%In Appendix \ref{adia_improve} we derive  a with the following simplifications:

@@

\item We consider eigenvalues such that
  $|\upsilon^2|\ll \kappa^2,\,\Omega^2$; which correspond to inertial
  waves \citepalias{barker15}. We call this the low frequency  
  approximation.

\item We are mainly interested in the effect of vertical shear, 
%  i.e. $q\neq
%  0$ for vertically isothermal disks. 
  which is already captured
  explicitly in the linearized equations through the term
  $d\Omega/dz$. We thus ignore the vertical dependence of $\Omega$ and
  $\kappa$ elsewhere, and replace them with
  $\Omega_K$.  We call this the Keplerian approximation.

@@

 In Appendix \ref{global_corr}  we discuss limitations of using
  the VGSB, where we show that this approximation is valid provided we 
  consider fast cooling times
  ($\beta\lesssim 1$) and perturbations with small radial lengthscales
  ($\khat\gtrsim 1$), as appropriate for the VSI.

### 2015-04-23 numerical.tex
fig updates, notation updates except for sec 3

However, as we have seen in Fig. \ref{bcrit_compare1}, the maximum
growth rate becomes insensitive to $\zmax$ for $\beta\gtrsim
\beta_\mathrm{crit}$. This suggests that the characteristic upper
limit on the thermal timescale for the VSI is not sensitive to
$\zmax$, so that vertically isothermal disks are in fact suitable for
the purpose of this paper.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gcorr_compare_vnoniso_maxrate} 
  \caption{Maximum VSI growth rates in the vertically non-isothermal disk
    model as a function of the thermal relaxation timescale
    $\beta$. The vertical line is a conjectured 
    critical thermal timescale given by
    $ h|s|/(\gamma-\Gamma)=0.158$.  
    \label{bcrit_compare2}}   
\end{figure}

>>>>

However, as we have seen in Fig. \ref{bcrit_compare1}, the maximum
growth rate becomes insensitive to $\zmax$ for $\beta\gtrsim
\beta_\mathrm{crit}$. This suggests that the characteristic upper
limit on the thermal timescale for the VSI is not sensitive to
$\zmax$, so that vertically isothermal disks are in fact suitable for
the purpose of characterizing the thermodynamic dependence of the
VSI.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gcorr_compare_vnoniso_maxrate} 
  \caption{Maximum VSI growth rates in the vertically non-isothermal disk
    model as a function of the thermal relaxation timescale
    $\beta$. The vertical dashed line is the critical thermal
    timescale given by Eq. \ref{iso_vsi_cond}. A conjectured critical thermal
    timescale accounting for a non-isothermal background, given by 
    $\beta_\mathrm{crit,noniso}= h|s|/(\gamma-\Gamma)=0.158$,  
    is also shown as the vertical dotted line. 
    \label{bcrit_compare2}}   
\end{figure}
